Draft -- Discriminating Seagrasses From Green Macroalgae in European Intertidal areas using high resolution multispectral drone imagery -- Draft

### 2024-11-13 index.qmd
adding pierres comments

  Coastal areas support seagrass meadows, which offer crucial ecosystem services including erosion control and carbon sequestration. However, these areas are increasingly impacted by human activities, leading to habitat fragmentation and seagrass decline. In situ surveys, traditionally performed to monitor these ecosystems face limitations on temporal and spatial coverage, particularly in intertidal zones, prompting the addition of satellite data within monitoring programs. Yet, satellite remote sensing can be limited by too coarse spatial and/or spectral resolution, making it difficult to discriminate seagrass from other macrophytes in highly heterogenous meadows. Drone (unmanned aerial vehicles – UAV) images at a very high spatial resolution offer a promising solution to address challenges related to spatial heterogeneity and intrapixel mixture. This study focuses on using drone acquisitions with a ten spectral band sensor similar to that onboard Sentinel-2, for mapping intertidal macrophytes at low tide (i.e. during a period of emersion) and effectively discriminating between seagrass and green macroalgae. Nine drone flights were conducted at two different altitudes (12 m and 120 m) across heterogeneous intertidal European habitats in France and Portugal, providing multispectral reflectance observation at very high spatial resolution (8 mm and 80 mm respectively). Taking advantage of their extremly high spatial resolution; the low altitude flights were used to train a Neural Network classifier to discrimintate five taxonomic classes of intertidal vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green macroalgae), Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae) and benthic Bacillariophyceae (Diatoms), and validated using concomitant field measurements. Classification of drone imagery resulted in an overall accuracy of 94% across all sites and images, covering a total area of 467 000 m². The model exhibited an accuracy of 96.4% in identifying seagrass. In particular , seagrass and green algae can be discriminated. The very high spatial resolution of the drone data made it possible to assess the influence of spatial resolution on the classification outputs, showing a limited loss in seagrass detection up to about 10m. Altogether, out findings suggest that the Multi-Spectral Instrument (MSI) onboard Sentinel-2 offers a relevant trade-off between its spatial and spectral resolution, thus offering promising perspectives for satellite remote sensing of intertidal biodiversity over lager scales. 

@@

At each location, a DJI Matrice 200 quadcopter drone equipped with a Micasense RedEdge Dual MX multispectral camera was flown to take 1.2 million pixel reflectance photographs with ten spectral bands ranging from the blue to the near infrared (NIR): 444, 475, 531, 560, 650, 668, 705, 717, 740 and 840 nm. To ensure consistent lighting conditions across flight paths, the drone's trajectory was aligned to maintain a solar azimuth angle of 90 degrees. An overlap of 70% and 80% (side and front respectively) between each image was set for each flight. A downwelling light sensor (DLS2) was used to acquire irradiance data concomitantly with the camera measurements. Raw data were calibrated in reflectance using a calibration panel reflective at ~50% provided by the manufacturer. Across all sites, flights were made at two different altitudes : 12 m or/and 120 m, with a spatial resolution of 8 mm and 80 mm, respectively (@tbl-flights). **Low-altitude flights, with a spatial resolution of 8 mm, were used to build the training dataset for the neural network, as this high resolution allowed for precise photo-interpretation of vegetation classes. In contrast, high-altitude flights were utilized for validation purposes.**

@@

Before each flight, targets used as ground control points were distributed over the study site and georeferenced with a Trimble © Geo XH 6000 differential GPS (dGPS). Ground control points were used to correct georeferencing imprecision of orthomosaics with an horizontal and vertical accuracy of 10cm. A dGPS was also used to georeference quadrats of 0.25 m², which assessed the presence or absence of five key taxonomic classes of intertidal vegetation: Bacillariophyceae **(unicellular benthic diatoms forming biofilms at the sediment surface during low tide, ranging in size from small patches (m²) to entire mudflats (km²); henceforth: Diatoms), Phaeophyceae (brown macroalgae that are generally attached to rocks or other substrates and often form dense beds in the intertidal zone; henceforth: brown macroalgae), Magnoliopsida (seagrasses, which are rooted flowering plants that grow submerged and form extensive meadows on soft sediments; henceforth: Seagrasses), Chlorophyceae (green macroalgae, typically found attached to rocks or drifting in the water, and known for their bright green color due to chlorophyll; henceforth: Green macroalgae), and Rhodophyceae (red macroalgae that are mostly attached to hard substrates and often occur in smaller patches; henceforth: red macroalgae)**. Only homogeneous vegetation patches extending over several meters were selected as ground control points. Pictures of each quadrat were uploaded online to the open-portal Global Biodiversity Information Facility (GBIF) platform [@BedeGbif]. Each photograph was also processed to estimate the percent cover of each type of vegetation using an image processing software [ImageJ, @schneider2012nih]. Hyperspectral reflectance signatures of each vegetation class were recorded using an ASD FieldSpec HandHeld 2 spectroradiometer, which acquires reflectance between 325 and 1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures served dual purposes: they validate the radiometric calibration of drone data and contribute to misclassification reduction in photo interpretations.

@@

```{r Figure2 Plot}
#| cache: true
#| echo: false
#| warning: false
#| fig-cap: "The five taxonomic classes of vegetation used to train the Neural Network model and an example of their raw spectral signatures at the spectral resolution of the Micasense RedEdge Dual MX. A : Magnoliopsida (*Zostera noltei*)  ; B : Phaeophyceae (*Fucus sp.*) ; C : Rhodophyceae (*Gracilaria vermiculophylla*) ; D : Chlorophyceae (*Ulva sp.*) ; E : Bacillariophyceae (Diatoms - MPB). **Classes were identified following the World register of Marine species.**"
#| label: fig-vegetation
#| out-width: "100%"

@@

```{r Figure3}
#| cache: false
#| echo: false
#| warning: false
#| fig-cap: "Schematic representation of the workflow. Parallelograms represent input or output data, and rectangles represent Python processing algorithms. The overall workflow of this study is divided into two distinct parts based on the spatial resolution of the drone flights: high-resolution flights (pixel size: 8 mm) were utilized for training and prediction of the Neural Network model, whereas lower-resolution flights (pixel size: 80 mm) were solely employed for prediction purposes. Validation has been performed on both high and low resolution flights."
#| label: fig-workflow
#| fig-width: 10
#| fig-height: 7
#| out-width: "100%"

@@

df_flx2<-df_training %>% 
  left_join(df_validation, by = "True") %>% 
  rename(`Validation Pixels` = "validation_pixels",
         `Taxonomic Class` = "True",
         `Training Pixels` = "Training_Pixels") %>% 
  mutate(`Validation Pixels` = case_when(`Taxonomic Class` == "Water" ~ `Validation Pixels`*4,
                                          T ~ `Validation Pixels`),
         `Validation Pixels` = case_when(`Taxonomic Class` == "Sediment" ~ round(`Validation Pixels`/4,0),
                                          T ~ `Validation Pixels`),
         Name = case_when( `Taxonomic Class` == "Bacillariophyceae" ~ "Diatoms",
                           `Taxonomic Class` == "Chlorophyceae" ~ "Green macroalgae",
                           `Taxonomic Class` == "Magnoliopsida" ~ "Seagrass",
                           `Taxonomic Class` == "Phaeophyceae" ~ "Brown macroalgae",
                          `Taxonomic Class` == "Rhodophyceae" ~ "Red macroalgae",
                          `Taxonomic Class` == "Water" ~ "Water",
                          `Taxonomic Class` == "Sediment" ~ "Sediment"),
         `Taxonomic Class` = case_when(`Taxonomic Class` == "Water" ~ "-",
                                        `Taxonomic Class` == "Sediment" ~ "-",
                                        T ~ `Taxonomic Class`)
         ) %>% 
  dplyr::select(Name,`Taxonomic Class`,`Training Pixels`,`Validation Pixels`)

@@

A dataset containing photo-interpreted drone reflectance pixels was built to train a Neural Network model. The training pixels were categorized into seven different classes, representing the various habitats encountered at the different study sites: Sediment, Water, green macroalgae, seagrasses, diatoms, brown macroalgae and red macroalgae. Only data from the low-altitude flights (@tbl-flights) were used for training because their 8 mm spatial resolution allowed to avoid spectral sub-pixel mixing and to accurately identify vegetation classes. **In the field, seagrasses displaying two distinct colors, each with a unique spectral signature, were observed. Careful attention was given to incorporating training pixels from both color types into the training dataset for the seagrass class. This approach was consistently applied to all classes within the model.** More than `r my_comma(round(sum(df_training$Training_Pixels),-3))` pixels at 8 mm resolution from the 3 training flights were used to train the model (@tbl-validationPX). Twenty one variables were used by the model as predictors: the ten raw spectral bands of the Micasense RedEdge Dual MX multispectral camera (ranging from 444 nm to 840 nm), the same ten spectral bands standardized using a min/max transformation ([@eq-std] ; @Cao2017) and the Normalized difference vegetation index (NDVI, @eq-ndvi). Standardisation of spectral bands is commonly used to eliminate the scaling differences between spectra and to limit the effect of biomass on the spectra shape [@Douay2022 ; @Davies2023].

@@

Variable Importance Plots (VIP) serve as a method to identify which predictors are important for predicting a specific class. Out of the 21 predictors utilized in this study, Variable Importance was computed only for the raw and standardized values of the 10 spectral bands captured by the MicaSense camera. This is achieved by repeatedly predicting the same dataset while randomly shuffling one predictor at a time. The benchmark score obtained after each iteration is then compared to the benchmark score obtained without shuffling any variables. The greater the difference between these two benchmark values, the more important the variable is for the model [@WEI2015399].

@@

We used a Generalized Linear Model (GLM) with a Beta distribution to examine the relationship between pixel resolution, vegetation class, and their interaction on the loss of vegetation. The loss of vegetation was modelled as function of the interaction between pixel resolution and vegetation class (Diatoms, brown macroalgae, seagrass, green macroalgae and red macroalage). Sample vs fitted residuals and quartile-quartile graphics were assessed visually, to ensure assumption of the models used were met.

@@

Each drone flight was used to produce a prediction map, as well as a probability map that indicates the model derived probability of the selected class for every pixel. The low-altitude flight conducted in Gafanha, Portugal, represented the site with the highest complexity (@fig-GafLow). Among the five vegetation classes on which the model was trained, four were present on this site, with green and red macroalgae mixed with a seagrass meadow. There was also diatoms forming biofilms on bare sediment surface. Although the seagrass was solely composed of a single specie, *Zostera noltei*, various colors of this specie could be observed from dark green (corresponding to healthy leaves) to brown (when leaves are senescent or have an altered pigment composition). Regardless of the variation of color, the class Magnoliopsida (seagrass) was accurately predicted by the model (F1 score of 0.96 at that site).

@@

A percent cover of at least 80% was sufficient to have all the 80 mm pixels correctly classified, with the exception of Magnoliopsida that required a higher cover (>90 %) to be accurately classified. Concerning the probability of each class, there is a linear relationship between the percent cover and the confidence of the model to predict the class. To predict green macroalgae with a model likelihood of 0.85, a cover of `r round((lm_green$coefficients[[2]]*85+lm_green$coefficients[[1]])*100,0)` % was needed, `r round((lm_magno$coefficients[[2]]*85+lm_magno$coefficients[[1]])*100,0)` % for seagrass, `r round((lm_Rodo$coefficients[[2]]*85+lm_Rodo$coefficients[[1]])*100,0)` % for red macroalgae and `r round((lm_MPB$coefficients[[2]]*85+lm_MPB$coefficients[[1]])*100,0)` % for diatoms When the vegetation cover of a given class was 100 %, coarser high-flight pixels were correctly classified for all the classes except for Bare Sediment, which was only correctly classified 80% of the time. This phenomenon may be attributed to the time gap between the two flights, allowing for microphytobenthos migration to the sediment surface during low tide, consequently altering the model's classification from bare sediment to Bacillariophyceae.

>>>>

  Coastal areas support seagrass meadows, which offer crucial ecosystem services including erosion control and carbon sequestration. However, these areas are increasingly impacted by human activities, leading to habitat fragmentation and seagrass decline. In situ surveys, traditionally performed to monitor these ecosystems face limitations on temporal and spatial coverage, particularly in intertidal zones, prompting the addition of satellite data within monitoring programs. Yet, satellite remote sensing can be limited by too coarse spatial and/or spectral resolution, making it difficult to discriminate seagrass from other macrophytes in highly heterogenous meadows. Drone (unmanned aerial vehicles – UAV) images at a very high spatial resolution offer a promising solution to address challenges related to spatial heterogeneity and intrapixel mixture. This study focuses on using drone acquisitions with a ten spectral band sensor similar to that onboard Sentinel-2, for mapping intertidal macrophytes at low tide (i.e. during a period of emersion) and effectively discriminating between seagrass and green macroalgae. Nine drone flights were conducted at two different altitudes (12 m and 120 m) across heterogeneous intertidal European habitats in France and Portugal, providing multispectral reflectance observation at very high spatial resolution (8 mm and 80 mm respectively). Taking advantage of their extremly high spatial resolution; the low altitude flights were used to train a Neural Network classifier to discrimintate five taxonomic classes of intertidal vegetation: Magnoliopsida (Seagrass), Chlorophyceae (Green macroalgae), Phaeophyceae (Brown algae), Rhodophyceae (Red macroalgae) and benthic Bacillariophyceae (Benthic diatoms), and validated using concomitant field measurements. Classification of drone imagery resulted in an overall accuracy of 94% across all sites and images, covering a total area of 467 000 m². The model exhibited an accuracy of 96.4% in identifying seagrass. In particular , seagrass and green algae can be discriminated. The very high spatial resolution of the drone data made it possible to assess the influence of spatial resolution on the classification outputs, showing a limited loss in seagrass detection up to about 10m. Altogether, out findings suggest that the Multi-Spectral Instrument (MSI) onboard Sentinel-2 offers a relevant trade-off between its spatial and spectral resolution, thus offering promising perspectives for satellite remote sensing of intertidal biodiversity over lager scales. 

@@

At each location, a DJI Matrice 200 quadcopter drone equipped with a Micasense RedEdge Dual MX multispectral camera was flown to take 1.2 million pixel reflectance photographs with ten spectral bands ranging from the blue to the near infrared (NIR): 444, 475, 531, 560, 650, 668, 705, 717, 740 and 840 nm. To ensure consistent lighting conditions across flight paths, the drone's trajectory was aligned to maintain a solar azimuth angle of 90 degrees. An overlap of 70% and 80% (side and front respectively) between each image was set for each flight. A downwelling light sensor (DLS2) was used to acquire irradiance data concomitantly with the camera measurements. Raw data were calibrated in reflectance using a calibration panel reflective at ~50% provided by the manufacturer. Across all sites, flights were made at two different altitudes : 12 m or/and 120 m, with a spatial resolution of 8 mm and 80 mm, respectively (@tbl-flights). **Low-altitude flights, with a spatial resolution of 8 mm, were used to build the training dataset for the neural network, as this high resolution allowed for precise photo-interpretation of vegetation classes. In contrast, high-altitude flights were used for validation purposes.**

@@

Before each flight, targets used as ground control points were distributed over the study site and georeferenced with a Trimble © Geo XH 6000 differential GPS (dGPS). Ground control points were used to correct georeferencing imprecision of orthomosaics with an horizontal and vertical accuracy of 10cm. A dGPS was also used to georeference quadrats of 0.25 m², which assessed the presence or absence of five key taxonomic classes of intertidal vegetation: Bacillariophyceae **(benthic diatoms forming biofilms at the sediment surface during low tide with biofilm's size ranging from small patches (m²) to entire mudflats (km²); henceforth: Benthic diatoms), Phaeophyceae (brown macroalgae generally attached to rocks or other substrates able to form dense beds in the intertidal zone; henceforth: Brown macroalgae), Magnoliopsida (seagrasses, rooted flowering marine plants able to form extensive meadows on soft sediments; henceforth: Seagrasses), Chlorophyceae (green macroalgae, typically found attached to rocks or washed ashore; henceforth: Green macroalgae), and Rhodophyceae (red macroalgae, attached to hard substrates and generally forming smaller patches; henceforth: Red macroalgae)**. Only homogeneous vegetation patches extending over several meters were selected as ground control points. Pictures of each quadrat were uploaded online to the open-portal Global Biodiversity Information Facility (GBIF) platform [@BedeGbif]. Each photograph was also processed to estimate the percent cover of each type of vegetation using an image processing software [ImageJ, @schneider2012nih]. Hyperspectral reflectance signatures of each vegetation class were recorded using an ASD FieldSpec HandHeld 2 spectroradiometer, which acquires reflectance between 325 and 1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures served dual purposes: they validate the radiometric calibration of drone data and contribute to misclassification reduction in photo interpretations.

@@

```{r Figure2 Plot}
#| cache: true
#| echo: false
#| warning: false
#| fig-cap: "The five taxonomic classes of vegetation used to train the Neural Network model and an example of their raw spectral signatures at the spectral resolution of the Micasense RedEdge Dual MX. A : Magnoliopsida (*Zostera noltei*)  ; B : Phaeophyceae (*Fucus sp.*) ; C : Rhodophyceae (*Gracilaria vermiculophylla*) ; D : Chlorophyceae (*Ulva sp.*) ; E : Bacillariophyceae (Benthic diatoms). **Classes were identified following the World register of Marine species.**"
#| label: fig-vegetation
#| out-width: "100%"

@@

```{r Figure3}
#| cache: false
#| echo: false
#| warning: false
#| fig-cap: "Schematic representation of the workflow. Parallelograms represent input or output data, and rectangles represent Python processing algorithms. The overall workflow of this study is divided into two distinct parts based on the spatial resolution of the drone flights: high-resolution flights (pixel size: 8 mm) were used for training and prediction of the Neural Network model, whereas lower-resolution flights (pixel size: 80 mm) were solely employed for prediction purposes. Validation has been performed on both high and low resolution flights."
#| label: fig-workflow
#| fig-width: 10
#| fig-height: 7
#| out-width: "100%"

@@

df_flx2<-df_training %>% 
  left_join(df_validation, by = "True") %>% 
  rename(`Validation Pixels` = "validation_pixels",
         `Taxonomic Class` = "True",
         `Training Pixels` = "Training_Pixels") %>% 
  mutate(`Validation Pixels` = case_when(`Taxonomic Class` == "Water" ~ `Validation Pixels`*4,
                                          T ~ `Validation Pixels`),
         `Validation Pixels` = case_when(`Taxonomic Class` == "Sediment" ~ round(`Validation Pixels`/4,0),
                                          T ~ `Validation Pixels`),
         Name = case_when( `Taxonomic Class` == "Bacillariophyceae" ~ "Benthic diatoms",
                           `Taxonomic Class` == "Chlorophyceae" ~ "Green macroalgae",
                           `Taxonomic Class` == "Magnoliopsida" ~ "Seagrass",
                           `Taxonomic Class` == "Phaeophyceae" ~ "Brown macroalgae",
                          `Taxonomic Class` == "Rhodophyceae" ~ "Red macroalgae",
                          `Taxonomic Class` == "Water" ~ "Water",
                          `Taxonomic Class` == "Sediment" ~ "Sediment"),
         `Taxonomic Class` = case_when(`Taxonomic Class` == "Water" ~ "-",
                                        `Taxonomic Class` == "Sediment" ~ "-",
                                        T ~ `Taxonomic Class`)
         ) %>% 
  dplyr::select(Name,`Taxonomic Class`,`Training Pixels`,`Validation Pixels`)

@@

A dataset containing photo-interpreted drone reflectance pixels was built to train a Neural Network model. The training pixels were categorized into seven different classes, representing the various habitats encountered at the different study sites: Sediment, Water, green macroalgae, seagrasses, Benthic diatoms, brown macroalgae and red macroalgae. Only data from the low-altitude flights (@tbl-flights) were used for training because their 8 mm spatial resolution allowed to avoid spectral sub-pixel mixing and to accurately identify vegetation classes. **In the field, seagrasses displaying two types of color were observed (i.e. whereas most seagrass had green leaves, brownish leaves were also observed due to senescence or photo-degradaration). Careful attention was given to incorporating training pixels from both color types into the training dataset for the seagrass class. This approach was consistently applied to all classes within the model.** More than `r my_comma(round(sum(df_training$Training_Pixels),-3))` pixels at 8 mm resolution from the 3 training flights were used to train the model (@tbl-validationPX). Twenty one variables were used by the model as predictors: the ten raw spectral bands of the Micasense RedEdge Dual MX multispectral camera (ranging from 444 nm to 840 nm), the same ten spectral bands standardized using a min/max transformation ([@eq-std] ; @Cao2017) and the Normalized difference vegetation index (NDVI, @eq-ndvi). Standardisation of spectral bands is commonly used to eliminate the scaling differences between spectra and to limit the effect of biomass on the spectra shape [@Douay2022 ; @Davies2023].

@@

Variable Importance Plots (VIP) serve as a method to identify which predictors are important for predicting a specific class. Out of the 21 predictors used in this study, Variable Importance was computed only for the raw and standardized values of the 10 spectral bands captured by the MicaSense camera. This is achieved by repeatedly predicting the same dataset while randomly shuffling one predictor at a time. The benchmark score obtained after each iteration is then compared to the benchmark score obtained without shuffling any variables. The greater the difference between these two benchmark values, the more important the variable is for the model [@WEI2015399].

@@

We used a Generalized Linear Model (GLM) with a Beta distribution to examine the relationship between pixel resolution, vegetation class, and their interaction on the loss of vegetation. The loss of vegetation was modelled as function of the interaction between pixel resolution and vegetation class (Benthic diatoms, brown macroalgae, seagrass, green macroalgae and red macroalage). Sample vs fitted residuals and quartile-quartile graphics were assessed visually, to ensure assumption of the models used were met.

@@

Each drone flight was used to produce a prediction map, as well as a probability map that indicates the model derived probability of the selected class for every pixel. The low-altitude flight conducted in Gafanha, Portugal, represented the site with the highest complexity (@fig-GafLow). Among the five vegetation classes on which the model was trained, four were present on this site, with green and red macroalgae mixed with a seagrass meadow. There was also benthic diatoms forming biofilms on bare sediment surface. Although the seagrass was solely composed of a single specie, *Zostera noltei*, various colors of this specie could be observed from dark green (corresponding to healthy leaves) to brown (when leaves are senescent or have an altered pigment composition). Regardless of the variation of color, the class Magnoliopsida (seagrass) was accurately predicted by the model (F1 score of 0.96 at that site).

@@

A percent cover of at least 80% was sufficient to have all the 80 mm pixels correctly classified, with the exception of Magnoliopsida that required a higher cover (>90 %) to be accurately classified. Concerning the probability of each class, there is a linear relationship between the percent cover and the confidence of the model to predict the class. To predict green macroalgae with a model likelihood of 0.85, a cover of `r round((lm_green$coefficients[[2]]*85+lm_green$coefficients[[1]])*100,0)` % was needed, `r round((lm_magno$coefficients[[2]]*85+lm_magno$coefficients[[1]])*100,0)` % for seagrass, `r round((lm_Rodo$coefficients[[2]]*85+lm_Rodo$coefficients[[1]])*100,0)` % for red macroalgae and `r round((lm_MPB$coefficients[[2]]*85+lm_MPB$coefficients[[1]])*100,0)` % for benthic diatoms When the vegetation cover of a given class was 100 %, coarser high-flight pixels were correctly classified for all the classes except for Bare Sediment, which was only correctly classified 80% of the time. This phenomenon may be attributed to the time gap between the two flights, allowing for microphytobenthos migration to the sediment surface during low tide, consequently altering the model's classification from bare sediment to Bacillariophyceae.
